---
title: "Quality of Life by Country:  A Clustering Analysis"
author: "Katherine M. Prioli"
date: "December 22, 2018"
output:
  pdf_document: default
  html_document: default
geometry: margin = 0.5in
references:
- URL: https://www.cdc.gov/nchs/products/databriefs/db328.htm
  author:
    - family: Murphy
      given: Sherry L.
    - family: Xu
      given: Jiaquan
    - family: Kochanek
      given: Kenneth D.
    - family: Arias
      given: Elizabeth
  id: cdc_lifeexp
  issued:
      year: 2018
  title: Mortality in the United States, 2017. NCHS Data Brief, no 328
  publisher: National Center for Health Statistics
- URL: https://github.com/kmprioliPROF/MAT_8790_Final_Project
  author:
  - family: Prioli
    given: Katherine M.
  id: kmpgit
  issued:
    month: 12
    year: 2018
  title: MAT_8790_Final_Project
- URL: http://apps.who.int/gho/data/view.main.SDG2016LEXv?lang=en
  author:
  - family: World Health Organization
  id: who_life
  issued:
    year: 2018
  title: Life Expectancy
- URL: http://apps.who.int/gho/data/view.main.182?lang=en
  author:
  - family: World Health Organization
  id: who_infantmort
  issued:
    year: 2018
  title: Probability of Dying per 1000 Live Births
- URL: http://reports.weforum.org/global-gender-gap-report-2016/rankings/
  author:
  - family: World Economic Forum
  id: wef_gender
  issued:
    year: 2016
  title: Gender Equality
- URL: http://reports.weforum.org/global-gender-gap-report-2016/measuring-the-global-gender-gap/
  author:
  - family: World Economic Forum
  id: wef_gender_desc
  issued:
    year: 2016
  title: Measuring the Global Gender Gap
- URL: https://data.worldbank.org/indicator/ny.gdp.mktp.cd?view=map&year_high_desc=true 
  author:
  - family: The World Bank
  id: worldbank_gdp
  issued:
    year: 2018
  title: Gross Domestic Product
- URL: http://hdr.undp.org/en/data
  author:
  - family: The United Nations Development Programme
  id: un_dvlpt
  issued:
    year: 2018
  title: Human Development Index
- URL: http://hdr.undp.org/en/content/human-development-index-hdi
  author:
  - family:  The United Nations Development Programme
  id: un_dvlpt_HDIdesc
  issued:
    year: 2018
  title: Human Development Reports - Human Development Index (HDI)
- URL: https://www.socialprogress.org/?tab=4
  author:
  - family: Social Progress Imperative
  id: socialprog
  issued:
    year: 2018
  title: Social Progress Index
- URL: https://www.socialprogress.org/index
  author:
  - family: Social Progress Imperative
  id: socialprog_desc
  issued:
    year: 2018
  title:  Social Progress Index - Overview
- URL: http://worldhappiness.report/ed/2018/
  author:
  - family: Helliwell
    given: John F.
  - family: Layard
    given: Richard
  - family: Sachs
    given: Jeffrey D.
  id: whr
  issued:
    year: 2018
  title: World Happiness Report
- URL: https://www-bcf.usc.edu/~gareth/ISL/
  author: 
  - family: James
    given: Gareth
  - family: Witten
    given: Daniela
  - family: Hastie
    given: Trevor
  - family: Tibshirani
    given: Robert
  id: intro_stat_learn
  issued:
    year: 2013
  title: An Introduction to Statistical Learning
  publisher: Springer
abstract: "Example abstract text"
---

<!-- Notes
  * Do a sensitivity analysis on cluster number (shiny?)
  * Clustering:  in Shiny app, write code to display "The United States is in " _fill in the blank_ "cluster"
  * Need to better organize flow of report (some results are in Methods, etc)
  * Need to more fully narrate steps and interpretations of output
  * Ensure uniform ordering of table contents, output, references
  
  * Limitations:
  

  -->


## **Background**

The Centers for Disease Control and Prevention (CDC) has recently issued a report indicating that life expectancy in the United States has decreased in 2017 as compared to 2016, with the overwhelming majority of deaths caused by heart disease and cancer, arguably preventable illnesses (@cdc_lifeexp).  Given that the United States has the world's largest economy, this decline in life expectancy is particularly concerning, and indicates that national wealth may not be predictive of citizens' longevity (@worldbank_gdp). 

As the world's economies trend toward globalism, there is increasing interest in understanding how these nations compare on key quality of life (QoL) factors, including but not limted to life expectancy.  Several organizations report on QoL measures as they evolve, including among others the World Economic Forum (WEF), World Health Organization (WHO), and the United Nations Development Programme.  The QoL measures reported by these bodies can be either unidimensional values or compound scores calculated from several factors of interest.

The objective of this analysis is to explore key QoL indicators by country, with particular focus on how the United States ranks, through a series of visualizations and *k*-means clustering analyses.


## **Methods**

This analysis included country-level QoL indicators as described in Table 1.

**Table 1.  Country-Level QoL measures.**

```{r table_1, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(kableExtra)

meas <- c("Gross Domestic Product (GDP)", "Infant mortality rate", "Life expectancy at birth", 
          "Life expectancy at sixty", "Human Development Index (HDI)", "Human Development Index (HDI)", 
          "Social Progress Index", "Global Gender Gap Index", "World Happiness Score")
sing_comp <- c(rep("Single", 4), rep("Compound", 5))
measdesc <- c("Valued in $US 2018", 
              "Number of infant deaths per 1,000 live births", 
              "Expected life at birth, both genders", 
              "Expected remaining life years at age 60, both genders", 
              "Developmental level, scale of 0:1", 
              "Developmental category, four levels (low, medium, high, very high)", 
              "Social progress level, scaled from 0:100 and comprising three broad categories:  basic human needs (e.g., nutrition, safety), foundations of wellbeing (e.g., basic knowledge, environmental quality), and opportunity (e.g., personal rights, freedoms)", 
              "Gender equality index, scaled from 0:1, based on measurements of gender-related gaps in such dimensions as economic participation, level of education, health and survival, and political offices held", 
              "Happiness score, scaled from 0:10, based on several factors including per-capita GDP, healthy life expectancy, social support, freedoms, and perception of corruption")
measref <- c("@worldbank_gdp", "@who_infantmort", "@who_life", "@who_life", "@un_dvlpt_HDIdesc", 
              "@un_dvlpt_HDIdesc", "@socialprog_desc", "@wef_gender_desc", "@whr")

table_1 <- as.tibble(cbind(meas, sing_comp, measdesc, measref))
colnames(table_1) <- c("Measure", "Single or Compound", "Description", "Source")
table_1_kbl <- table_1 %>% 
  kable(format = "latex") %>% 
  column_spec(1, width = "10em", border_left = FALSE, border_right = FALSE) %>% 
  column_spec(2, width = "6em", border_left = FALSE, border_right = FALSE) %>% 
  column_spec(3, width = "25em", border_left = FALSE, border_right = FALSE) %>% 
  column_spec(4, width = "8em", border_left = FALSE, border_right = FALSE)
table_1_kbl
```


Data for these measures was obtained for calendar year 2016 in `.csv` or `.xls(x)` formats.  Additionally, dataframe containing country identifiers (full names and three-letter codes) was generated from the `countrycode` library to facilitate merging the datafiles into one dataframe.

### *Wrangling and Exploration*

Each country-level datafile was imported, wrangled as needed, then tested against the dataframe containing country identifiers via `anti_join()` to identify mismatches.  Mismatching country names were manually recoded for each datafile, then all datafiles were merged using serial `lef_join()` statements.  Countries with wholly missing data were excluded.  The resulting dataframe, titled `alldata`, is presented in Table 2.

**Table 2.  `alldata` dataframe contents.**

| **Source**       	              | **Variable Name**    	| **Description**                                  	|
|-------------------------------- |----------------------	|--------------------------------------------------	|
| @socialprog      	              | SPI                  	| Social Progress Index value (scale of 0:100)     	|
| @worldbank_gdp   	              | GDP_USD_2018         	| 2016 Gross Domestic Product (valued in $US 2018) 	|
| @un_dvlpt        	              | HDIrank              	| Human Development Index ranking                  	|
| @un_dvlpt        	              | HDIindex             	| HDI index value (scale of 0:1)                   	|
| @un_dvlpt        	              | HDI_cat              	| HDI index category (5 levels)                    	|
| @whr             	              | happiness            	| World Happiness Score (scale of 0:10)            	|
| @wef_gender      	              | genderequality_index 	| Gender Equality Index (scale of 0:1)             	|
| @who_infantmort  	              | infantmort           	| Infant mortality rate                            	|
| @who_life        	              | birth_MF             	| Life expectancy at birth, males & females        	|
| @who_life        	              | sixty_MF             	| Life expectancy at 60 years, males & females     	|

At least one univariate visualization was generated for each variable in `alldata` via `ggplot()`, and a correlation matrix was produced to investigate pairwise relationshps between continuous variables.  Next, a series of ordered country-as-factor bivariate visualizations were created to explore the top and bottom 20 countries by ranking within each variable.

### *k-Means Clustering*

To assess how the United States compares to the rest of the world, a series of *k*-means cluster analyses was carried out.  *K*-means clustering is described elsewhere; briefly, given bivariate data and a desired number *k* groups, this classification algorithm classifies the points in the two-dimensional plane to minimize the total within-cluster variation for all clusters (@intro_stat_learn).  This is an iterative process that works by establishing *k* centroids, classifying each point by which centroid is closest, then moving the centroids to the center of their corrresponding clusters, and repeating the process.  Iteration terminates when the centroids no longer move, and the classification established in this terminal iteraton is the clustering.

The Human Development Index categorizes the world's countries into four developmental levels (low, medium, high, and very high); thus *k*-means clustering analysis was performed assuming 4 clusters.  Missing values were excluded to ensure the clustering algorithm would run, and a function was written to subset the clustering dataset (named `clusterdata`) to the variables of interest for each *k*-means analysis.  On each output plot, the United States was identified by an enlarged `geom_point()`.

For brevity, example code for wrangling, plotting, and clustering is presented in the Results section; full code is available in the GitHub repository for this report (@kmpgit).  The libraries required for this analysis were loaded as shown below.

```{r setup, include = TRUE, warning = FALSE, results = 'hide', echo = TRUE, message = FALSE}
library(tidyverse)
library(readxl)          # For importing .xls(x) datasets
library(lazyeval)        # For renaming columns in function
library(countrycode)     # For establishing uniform country identifiers
library(ggthemr)         # For prettifying output
library(gridExtra)       # For grid.arrange()
library(grid)            # For textGrob() to annotate grid.arrange() elements
library(kableExtra)      # For nicer output tables
library(GGally)          # For ggpairs() correlation matrix

ggthemr("fresh")
```


### *Example code*

Establishing a crosswalk for country names and 3-letter codes:

```{r countries, include = TRUE, echo = TRUE, eval = TRUE}
countries_full <- codelist_panel %>% 
  select(country.name.en, year, genc3c, iso3c, wb_api3c) %>% 
  group_by(country.name.en) %>% 
  mutate(maxyr = max(year)) %>% 
  ungroup %>% 
  mutate(maxyr = case_when(
    maxyr == year ~ 1,
    TRUE ~ 0
  )) %>% 
  filter(maxyr == 1) %>% 
  select(-maxyr) %>% 
  distinct()

countries_full <- countries_full %>% 
  mutate(country3 = case_when(
    iso3c == genc3c & iso3c == wb_api3c ~ iso3c,
    is.na(iso3c) == FALSE ~ iso3c,
    is.na(iso3c) == TRUE & is.na(genc3c) == FALSE ~ genc3c,
    is.na(iso3c) == TRUE & is.na(genc3c) == TRUE & is.na(wb_api3c) == FALSE ~ wb_api3c
  )) %>% 
  rename(country = country.name.en) %>% 
  arrange(country)
  
countries <- countries_full %>% 
  select(country, country3)
```

Importing and wrangling each data file, and standardizing country names:

```{r HDI_import, include = TRUE, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE}

# Importing raw data

HDIraw <- read_xlsx("data/HDIdata2018.xlsx", sheet = "Table 2")

# Selecting columns of interest

HDIdata <- HDIraw %>%
  select(1:2, X__14)

# Assigning sensible column names

HDIcolnm <- c(HDIdata[[3,1]], HDIdata[[3,2]], HDIdata[[4,3]])
colnames(HDIdata) <- HDIcolnm

# Determining boundaries for human development levels in the data
# and using these to create one dataframe for each level

vhhd_st <- which(HDIdata$Country == "VERY HIGH HUMAN DEVELOPMENT") + 1
vhhd_end <- which(HDIdata$Country == "HIGH HUMAN DEVELOPMENT") - 1

hhd_st <- which(HDIdata$Country == "HIGH HUMAN DEVELOPMENT") + 1
hhd_end <- which(HDIdata$Country == "MEDIUM HUMAN DEVELOPMENT") - 1

mhd_st <- which(HDIdata$Country == "MEDIUM HUMAN DEVELOPMENT") + 1
mhd_end <- which(HDIdata$Country == "LOW HUMAN DEVELOPMENT") - 1

lhd_st <- which(HDIdata$Country == "LOW HUMAN DEVELOPMENT") + 1
lhd_end <- which(HDIdata$Country == "OTHER COUNTRIES OR TERRITORIES") - 1

oth_st <- which(HDIdata$Country == "OTHER COUNTRIES OR TERRITORIES") + 1
oth_end <- which(HDIdata$Country == "Human development groups") - 2

HDI_vhhd <- HDIdata %>% 
  slice(vhhd_st:vhhd_end) %>% 
  mutate(HDI_cat = "Very High")

HDI_hhd <- HDIdata %>% 
  slice(hhd_st:hhd_end) %>% 
  mutate(HDI_cat = "High")

HDI_mhd <- HDIdata %>% 
  slice(mhd_st:mhd_end) %>% 
  mutate(HDI_cat = "Medium")

HDI_lhd <- HDIdata %>% 
  slice(lhd_st:lhd_end) %>% 
  mutate(HDI_cat = "Low")

HDI_oth <- HDIdata %>% 
  slice(oth_st:oth_end) %>% 
  mutate(HDI_cat = NA)

# Combining the dataframes into one 

HDIdata <- bind_rows(HDI_vhhd, HDI_hhd, HDI_mhd, HDI_lhd, HDI_oth) %>% 
  rename(HDIrank = `HDI rank`) %>% 
  rename(country = Country) %>% 
  rename(HDIindex = `2016`) %>% 
  mutate(HDI_cat = factor(HDI_cat, levels = c("Low", "Medium", "High", "Very High"))) %>% 
  mutate(HDIrank = case_when(
    HDIrank == ".." ~ as.numeric(NA),
    TRUE ~ as.numeric(HDIrank)
  )) %>% 
  mutate(HDIindex = case_when(
    HDIindex == ".." ~ as.numeric(NA),
    TRUE ~ as.numeric(HDIindex)
  ))
HDIdata <- HDIdata[c(2, 1, 3:4)]
```

Standardizing country names by using `anti_join()` to see which countries in `HDIdata` don't have a match in the `countries` dataframe, and correcting those for which an inexact match exists:

```{r HDI_countrytest, include = TRUE, echo = TRUE, eval = TRUE}
HDIanti <- HDIdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country)
dim(HDIanti)
```

Correcting for mismatches with `countries` using `mutate()`:

```{r HDI_countryfix, include = TRUE, echo = TRUE, eval = TRUE}

HDIdata <- HDIdata %>%
  mutate(country = case_when(
    country == "Antigua and Barbuda"                       ~ "Antigua & Barbuda",
    country == "Bolivia (Plurinational State of)"          ~ "Bolivia",
    country == "Bosnia and Herzegovina"                    ~ "Bosnia & Herzegovina",
    country == "Brunei Darussalam"                         ~ "Brunei",
    country == "Cabo Verde"                                ~ "Cape Verde",
    country == "Congo"                                     ~ "Congo - Brazzaville",
    country == "Congo (Democratic Republic of the)"        ~ "Congo - Kinshasa",
    country == "Eswatini (Kingdom of)"                     ~ "Swaziland",
    country == "Hong Kong, China (SAR)"                    ~ "Hong Kong SAR China",
    country == "Iran (Islamic Republic of)"                ~ "Iran",
    country == "Korea (Democratic People's Rep. of)"       ~ "North Korea",
    country == "Korea (Republic of)"                       ~ "South Korea",
    country == "Lao People's Democratic Republic"          ~ "Laos",
    country == "Moldova (Republic of)"                     ~ "Moldova",
    country == "Myanmar"                                   ~ "Myanmar (Burma)",
    country == "Palestine, State of"                       ~ "Palestinian Territories",
    country == "Russian Federation"                        ~ "Russia",
    country == "Saint Kitts and Nevis"                     ~ "St. Kitts & Nevis",
    country == "Saint Lucia"                               ~ "St. Lucia",
    country == "Saint Vincent and the Grenadines"          ~ "St. Vincent & Grenadines",
    country == "Syrian Arab Republic"                      ~ "Syria",
    country == "Tanzania (United Republic of)"             ~ "Tanzania",
    country == "The former Yugoslav Republic of Macedonia" ~ "Macedonia",
    country == "Trinidad and Tobago"                       ~ "Trinidad & Tobago",
    country == "Venezuela (Bolivarian Republic of)"        ~ "Venezuela",
    country == "Viet Nam"                                  ~ "Vietnam",
    country == "C么te d'Ivoire"                             ~ as.character(NA),   # UTC-8
    country == "Sao Tome and Principe"                     ~ as.character(NA),   # conflicts
    TRUE                                                   ~ as.character(country)
  )) %>% 
  filter(!is.na(country))

HDIanti <- HDIdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country)
dim(HDIanti)
```

```{r HDI_cleanup, include = FALSE, echo = FALSE, eval = TRUE}
rm(list = c("hhd_end", "hhd_st", "lhd_end", "lhd_st", "mhd_end", "mhd_st", "oth_end", "oth_st", "vhhd_end",
            "vhhd_st", "HDIraw", "HDIcolnm", "HDI_vhhd", "HDI_hhd", "HDI_mhd", "HDI_lhd", "HDI_oth", "HDIanti"))
```

<!-- Social Progress Index (SPI) data: -->

```{r SPI_import, include = FALSE, echo = FALSE, eval = TRUE}
SPI_2016_raw <- read_xlsx("data/SPIdata.xlsx", sheet = 4)

SPIdata <- SPI_2016_raw %>% 
  select(2:3) %>% 
  rename(`SPI` = `Social Progress Index`,
         country3 = Code)
```

<!-- Testing against `countries` and renaming to correct for conflicts: -->

```{r SPI_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
SPIanti <- SPIdata %>%
  anti_join(countries, by = "country3") %>%
  select(country3) %>%
  arrange(country3) %>% 
  unique()
dim(SPIanti)

SPIdata <- SPIdata %>%
  mutate(country3 = case_when(
    country3 == "CHI" ~ as.character(NA), # Nonstandard code for Chile; omitting (no data in these rows)
    country3 == "KSV" ~ "XKS",            # Nonstandard code for Kosovo
    country3 == "NCY" ~ as.character(NA), # Turk. Repub. of N. Cyprus; omitting (conflict w/Cyprus)
    country3 == "SML" ~ as.character(NA), # Unable to determine
    country3 == "WBG" ~ as.character(NA), # West Bank / Gaza Strip; omitting (conflict w/Palestine)
    TRUE ~ as.character(country3)
  )) %>% 
  filter(!is.na(country3))

SPIanti <- SPIdata %>%
  anti_join(countries, by = "country3") %>%
  select(country3) %>%
  arrange(country3) %>% 
  unique()
dim(SPIanti)

# Removing unneeded files

rm(list = c("SPI_2016_raw", "SPIanti"))
```

<!-- World Happiness Score data: -->

```{r WHRdata_import, include = FALSE, echo = FALSE, eval = TRUE}
WHR_raw <- read_xls("data/WHRdata.xls", sheet = 1)
WHRdata <- WHR_raw %>% 
  select(1:3) %>% 
  rename(happiness = `Life Ladder`) %>% 
  filter(year == 2016) %>% 
  select(-year)
```

<!-- Testing against `countries` and correcting for conflicts: -->

```{r WHR_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
WHRanti <- WHRdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(WHRanti)

WHRdata <- WHRdata %>% 
  mutate(country = case_when(
    country == "Bosnia and Herzegovina"                    ~ "Bosnia & Herzegovina",
    country == "Congo (Brazzaville)"                       ~ "Congo - Brazzaville",
    country == "Congo (Kinshasa)"                          ~ "Congo - Kinshasa",
    country == "Czech Republic"                            ~ "Czechia",
    country == "Hong Kong S.A.R. of China"                 ~ "Hong Kong SAR China",
    country == "Myanmar"                                   ~ "Myanmar (Burma)",
    country == "Taiwan Province of China"                  ~ "Taiwan",
    country == "Trinidad and Tobago"                       ~ "Trinidad & Tobago",
    country == "North Cyprus"                              ~ as.character(NA),   # conflicts w/Cyprus
    country == "Ivory Coast"                               ~ as.character(NA),   # UTC-8 conflicts;
    TRUE                                                   ~ as.character(country)
  )) %>% 
  filter(!is.na(country))

WHRanti <- WHRdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(WHRanti)

# Removing unneeded files

rm(list = c("WHR_raw", "WHRanti"))
```

<!-- World Economic Forum Gender Equality Index data: -->

```{r gender_import, include = FALSE, echo = FALSE, eval = TRUE}
gender_raw <- read_xlsx("data/WEF_gender.xlsx", sheet = 2, range = "$A$2:$K$147")
gender_raw <- gender_raw %>% 
  select(1:2)
colnames(gender_raw) <- c("country", "genderequality_index")
genderdata <- gender_raw %>% 
  slice(2:dim(gender_raw)[1]) %>% 
  mutate(genderequality_index = as.numeric(genderequality_index))
```

<!-- Testing against `countries` and correcting for conflicts: -->

```{r gender_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
genderanti <- genderdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(genderanti)

genderdata <- genderdata %>%
  mutate(country = case_when(
    country == "Bosnia and Herzegovina"                    ~ "Bosnia & Herzegovina",
    country == "Brunei Darussalam"                         ~ "Brunei",
    country == "Czech Republic"                            ~ "Czechia",
    country == "Gambia, The"                               ~ "Gambia",
    country == "Iran, Islamic Rep."                        ~ "Iran",
    country == "Korea, Rep."                               ~ "South Korea",
    country == "Kyrgyz Republic"                           ~ "Kyrgyzstan",
    country == "Lao PDR"                                   ~ "Laos",
    country == "Macedonia, FYR"                            ~ "Macedonia",
    country == "Russian Federation"                        ~ "Russia",
    country == "Slovak Republic"                           ~ "Slovakia",
    country == "Trinidad and Tobago"                       ~ "Trinidad & Tobago",
    country == "C么te d'Ivoire"                             ~ as.character(NA),   # UTC-8 conflict
    TRUE                                                   ~ as.character(country)
  )) %>% 
  filter(!is.na(country))

genderanti <- genderdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(genderanti)

# Removing unneeded files

rm(list = c("gender_raw", "genderanti"))
```

<!-- World Health Organization infant mortality rate data: -->

```{r infantmort_import, include = FALSE, echo = FALSE, eval = TRUE}
infantmort_raw <- read_csv("data/WHO_infantmort.csv")
infantmortdata <- infantmort_raw %>% 
  select(1, 3) %>% 
  rename(country = Country) %>% 
  rename(infantmort = `Infant mortality rate (probability of dying between birth and age 1 per 1000 live births)`) %>% 
  mutate(infantmort = as.numeric(infantmort) / 1000)
```

<!-- Testing against `countries` and correcting for conflicts: -->

```{r infantmort_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
infantmortanti <- infantmortdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(infantmortanti)

infantmortdata <- infantmortdata %>% 
  mutate(country = case_when(
    country == "Antigua and Barbuda"                                  ~ "Antigua & Barbuda",
    country == "Bolivia (Plurinational State of)"                     ~ "Bolivia",
    country == "Bosnia and Herzegovina"                               ~ "Bosnia & Herzegovina",
    country == "Brunei Darussalam"                                    ~ "Brunei",
    country == "Cabo Verde"                                           ~ "Cape Verde",
    country == "Congo"                                                ~ "Congo - Brazzaville",
    country == "Democratic People's Republic of Korea"                ~ "North Korea",
    country == "Democratic Republic of the Congo"                     ~ "Congo - Kinshasa",
    country == "Eswatini"                                             ~ "Swaziland",
    country == "Iran (Islamic Republic of)"                           ~ "Iran",
    country == "Lao People's Democratic Republic"                     ~ "Laos",
    country == "Myanmar"                                              ~ "Myanmar (Burma)",
    country == "Republic of Korea"                                    ~ "South Korea",
    country == "Republic of Moldova"                                  ~ "Moldova",
    country == "Russian Federation"                                   ~ "Russia",
    country == "Saint Kitts and Nevis"                                ~ "St. Kitts & Nevis",
    country == "Saint Lucia"                                          ~ "St. Lucia",
    country == "Saint Vincent and the Grenadines"                     ~ "St. Vincent & Grenadines",
    country == "Syrian Arab Republic"                                 ~ "Syria",
    country == "The former Yugoslav republic of Macedonia"            ~ "Macedonia",
    country == "Trinidad and Tobago"                                  ~ "Trinidad & Tobago",
    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
    country == "United Republic of Tanzania"                          ~ "Tanzania",
    country == "United States of America"                             ~ "United States",
    country == "Venezuela (Bolivarian Republic of)"                   ~ "Venezuela",
    country == "Viet Nam"                                             ~ "Vietnam",
    country == "C么te d'Ivoire"                                        ~ as.character(NA),   # UTC-8
    country == "Sao Tome and Principe"                                ~ as.character(NA),   # conflicts
    TRUE                                                              ~ as.character(country)
  )) %>% 
  filter(!is.na(country))

infantmortanti <- infantmortdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(infantmortanti)

rm(list = c("infantmort_raw", "infantmortanti"))
```

<!-- World Health Organization life expectancy data: -->

```{r lifeexp_import, include = FALSE, echo = FALSE, eval = TRUE}
lifeexp_raw <- read_csv("data/WHO_lifeexp.csv")
lifeexp_raw <- lifeexp_raw %>% select(1:3, 6)

lifeexp_sub <- lifeexp_raw

colnames(lifeexp_sub) <- c("country", "year", "birth_MF", "sixty_MF")
lifeexpdata <- lifeexp_sub %>% 
  slice(2:length(country)) %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(birth_MF = as.numeric(birth_MF)) %>% 
  mutate(sixty_MF = as.numeric(sixty_MF)) %>% 
  filter(year == 2016) %>% 
  select(-year)
```

<!-- Testing against `countries` and correcting for conflicts: -->

```{r lifeexp_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
lifeexpanti <- lifeexpdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(lifeexpanti)

lifeexpdata <- lifeexpdata %>% 
  mutate(country = case_when(
    country == "Antigua and Barbuda"                                  ~ "Antigua & Barbuda",
    country == "Bolivia (Plurinational State of)"                     ~ "Bolivia",
    country == "Bosnia and Herzegovina"                               ~ "Bosnia & Herzegovina",
    country == "Brunei Darussalam"                                    ~ "Brunei",
    country == "Cabo Verde"                                           ~ "Cape Verde",
    country == "Congo"                                                ~ "Congo - Brazzaville",
    country == "Democratic People's Republic of Korea"                ~ "North Korea",
    country == "Democratic Republic of the Congo"                     ~ "Congo - Kinshasa",
    country == "Eswatini"                                             ~ "Swaziland",
    country == "Iran (Islamic Republic of)"                           ~ "Iran",
    country == "Lao People's Democratic Republic"                     ~ "Laos",
    country == "Myanmar"                                              ~ "Myanmar (Burma)",
    country == "Republic of Korea"                                    ~ "South Korea",
    country == "Republic of Moldova"                                  ~ "Moldova",
    country == "Russian Federation"                                   ~ "Russia",
    country == "Saint Lucia"                                          ~ "St. Lucia",
    country == "Saint Vincent and the Grenadines"                     ~ "St. Vincent & Grenadines",
    country == "Syrian Arab Republic"                                 ~ "Syria",
    country == "The former Yugoslav republic of Macedonia"            ~ "Macedonia",
    country == "Trinidad and Tobago"                                  ~ "Trinidad & Tobago",
    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
    country == "United Republic of Tanzania"                          ~ "Tanzania",
    country == "United States of America"                             ~ "United States",
    country == "Venezuela (Bolivarian Republic of)"                   ~ "Venezuela",
    country == "Viet Nam"                                             ~ "Vietnam",
    country == "C么te d'Ivoire"                                        ~ as.character(NA),   # UTC-8
    country == "Sao Tome and Principe"                                ~ as.character(NA),   # conflicts
    TRUE                                                              ~ as.character(country)
  )) %>% 
  filter(!is.na(country))

lifeexpanti <- lifeexpdata %>%
  anti_join(countries, by = "country") %>%
  select(country) %>%
  arrange(country) %>% 
  unique()
dim(lifeexpanti)

# Removing unneeded files

rm(list = c("lifeexp_raw", "lifeexp_sub"))
```

<!-- World Bank Gross Domestic Product (GDP) data: -->

```{r GDP_import, include = FALSE, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
GDPraw <- read_csv("data/GDPdata.csv")

GDPcolnames <- GDPraw %>% slice(4:4)
GDPsub <- GDPraw %>% 
  slice(-(1:4))
colnames(GDPsub) <- GDPcolnames

GDPdata <- GDPsub %>% 
  rename(country = `Country Name`) %>% 
  rename(country3 = `Country Code`) %>% 
  select(-(3:44)) %>%
  gather(3:20, key = "year", value = "GDP_USD_2018") %>%      # Valuation is in $US 2018
  mutate(year = as.numeric(year)) %>% 
  filter(year == 2016) %>% 
  select(-year)
```

<!-- Testing against `countries` and correcting for conflicts: -->

```{r GDP_countrytest, include = FALSE, echo = FALSE, eval = TRUE}
GDPanti <- GDPdata %>%
  anti_join(countries, by = "country3") %>%
  select(country3, country) %>%
  arrange(country3, country) %>% 
  unique()
dim(GDPanti)

GDPdata <- GDPdata %>% 
  mutate(country3 = case_when(
    country3 == "XKX" ~ "XKS",
    country3 == "CHI" ~ as.character(NA), # Channel Islands
    country3 %in% c("ARB", "CEB", "CSS", "EAP", "EAR", "EAS", "ECA", "ECS", "EMU",  # Region-of-world codes
                    "EUU", "FCS", "HIC", "HPC", "IBD", "IBT", "IDA", "IDB", "IDX",  # (not country codes)
                    "INX", "LAC", "LCN", "LDC", "LIC", "LMC", "LMY", "LTE", "MEA", 
                    "MIC", "MNA", "NAC", "OED", "OSS", "PRE", "PSS", "PST", "SAS", 
                    "SSA", "SSF", "SST", "TEA", "TEC", "TLA", "TMN", "TSA", "TSS", 
                    "UMC", "WLD") ~ as.character(NA),
    TRUE ~ as.character(country3)
  )) %>% 
  filter(!is.na(country3)) %>% 
  select(-country)

GDPanti <- GDPdata %>%
  anti_join(countries, by = "country3") %>%
  select(country3) %>%
  arrange(country3) %>% 
  unique()
dim(GDPanti)

# Removing unneeded files

rm(list = c("GDPraw", "GDPsub", "GDPcolnames", "GDPanti"))
```

Combining individual data files into one dataframe and filtering out countries with no data:

```{r combine, include = TRUE, echo = TRUE, eval = TRUE}
joindata_1 <- full_join(countries, HDIdata, by = "country")
joindata_2 <- left_join(joindata_1, SPIdata, by = "country3")
joindata_3 <- left_join(joindata_2, WHRdata, by = "country")
joindata_4 <- left_join(joindata_3, genderdata, by = "country")
joindata_5 <- left_join(joindata_4, infantmortdata, by = "country")
joindata_6 <- left_join(joindata_5, lifeexpdata, by = "country")
joindata_7 <- left_join(joindata_6, GDPdata, by = "country3")

joinsub <- joindata_7 %>% 
  arrange(country) %>% 
  mutate(exclude_flag = case_when(
    is.na(HDIrank) == TRUE & 
      is.na(HDIindex) == TRUE &
      is.na(HDI_cat) == TRUE &
      is.na(SPI) == TRUE &
      is.na(happiness) == TRUE &
      is.na(genderequality_index) == TRUE &
      is.na(infantmort) == TRUE &
      is.na(birth_MF) == TRUE &
      is.na(sixty_MF) == TRUE &
      is.na(GDP_USD_2018) == TRUE             ~ TRUE,
    TRUE                                      ~ FALSE
  )) %>% 
  filter(exclude_flag == FALSE) %>% 
  select(-exclude_flag)

alldata <- joinsub %>% 
  mutate(country = factor(country)) %>% 
  mutate(country3 = factor(country3)) %>% 
  mutate(US = case_when(
    country == "United States" ~ "US",
    TRUE                       ~ "Non US"
  )) %>% 
  mutate(color = case_when(
    country == "United States" ~ "red",
    TRUE                       ~ "#545454"
  ))

alldata <- alldata[c(1:2, 13:14, 6, 12, 3:5, 7:11)]

len <- dim(alldata)[[1]]

# write_csv(alldata, paste0("data/alldata_", lubridate::today(),".csv"))  # Uncomment to export data
```

```{r join_remove, include = FALSE, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
# Removing unneeded files

rm(list = c("joindata_1", "joindata_2", "joindata_3", "joindata_4",
            "joindata_5", "joindata_6", "joindata_7", "joinsub"))
```

Example univariate visualization code:

```{r SPIplot_code, include = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
SPI_hist <- ggplot(data = alldata, aes(x = SPI)) +
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$SPI))))) +
  xlab("Social Progress Index") +
  ylab("Count") +
  ggtitle("Social Progress Index Distribution")
```

Example ordered country-level plot code:

```{r SPI_country, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 6}
alldata_SPI <- alldata %>%
  filter(!is.na(SPI) == TRUE) %>% 
  arrange(desc(SPI)) %>% 
  select(SPI, country, US, color)

alldata_SPI_top20 <- alldata_SPI %>% head(20)
alldata_SPI_bot20 <- alldata_SPI %>% tail(20)
alldata_SPI40 <- bind_rows(alldata_SPI_top20, alldata_SPI_bot20)

colors <- alldata_SPI40$color[order(alldata_SPI40$SPI)]

SPI_country_point <- ggplot(data = alldata_SPI40, aes(x = SPI, 
                                                      y = fct_reorder(country, SPI), color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 100) +
  xlab("SPI") +
  ylab("Country") +
  ggtitle("Social Progress Index by Country, \nTop and Bottom 20 Countries")
```

## **Results**

### *Data Exploration and Visualizations*

First, exploring the Social Progress Index data:

```{r SPIplot, include = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
SPI_hist
```

Next, exploring GDP by summary statistics:

```{r GDPsumm, include = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
GDPsumm <- broom::tidy(round(summary(alldata$GDP_USD_2018 / 1000000), digits = 4)) %>% 
  kable(format = "markdown")
GDPsumm
```

Taking the log transform and plotting:

```{r GDPplot, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
GDP_hist <- ggplot(data = alldata, aes(x = log(GDP_USD_2018))) +
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$GDP_USD_2018))))) +
  xlab("log(GDP) ($US 2018)") +
  ylab("Count") +
  ggtitle("Gross Domestic Product Distribution, Log Transform") +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$"))
GDP_hist
```

Exploring the Human Development Index variables:

```{r HDIplots, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 4}
HDIindex_hist <- ggplot(data = alldata, aes(x = HDIindex)) +
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$HDIindex))))) +
  xlab("Human Development Index") +
  ylab("Count") +
  ggtitle("Human Development Index Distribution")

HDIcat_bar <- ggplot(data = alldata, aes(x = HDI_cat)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), color = "#233b43",
            hjust = 0.5, position = position_stack(vjust = 0.5)) +
  xlab("Human Development Category") +
  ylab("Count") +
  ggtitle("Human Development Index Counts \nby Category")

grid.arrange(HDIindex_hist, HDIcat_bar, nrow = 1)
```

Exploring the World Happiness Report data:

```{r happinessplot, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
happiness_hist <- ggplot(data = alldata, aes(happiness)) +
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$happiness))))) +
  xlab("Happiness Score") +
  ylab("Count") +
  ggtitle("Happiness Score Distribution")
happiness_hist
```

Exploring the gender equality index data:

```{r genderplot, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
gender_hist <- ggplot(data = alldata, aes(x = genderequality_index)) +
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$genderequality_index))))) +
  xlab("Gender Equality Index") +
  ylab("Count") +
  ggtitle("Gender Equality Index Distribution")
gender_hist
```

Exploring the WHO infant mortality rate data:

```{r infantmortplot, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4}
infantmort_hist <- ggplot(data = alldata, aes(x = infantmort)) + 
  geom_histogram(bins = ceiling(sqrt(len - sum(is.na(alldata$genderequality_index))))) +
  xlab("Probability of Infant Mortality") +
  ylab("Count") +
  ggtitle("Infant Mortality Distribution")
infantmort_hist
```

Exploring the WHO life expectancy data:

```{r lifeexpplot, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 4}
lifeexp_birth_box <- ggplot(data = alldata, aes(y = birth_MF)) +
  geom_boxplot() +
  ylim(50, 100) +
  ylab("Total Life Expectancy (years)") +
  ggtitle("...at Birth") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

lifeexp_sixty_box <- ggplot(data = alldata, aes(y = 60 + sixty_MF)) +
  geom_boxplot() +
  ylim(50, 100) +
  ylab("") +
  ggtitle("...at Age 60") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y.left = element_blank())

grid.arrange(lifeexp_birth_box, lifeexp_sixty_box, nrow = 1,
             top = textGrob("Total Life Expectancy...",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#545454")))
```

Investigating pairwise relationships between continuous variables:

```{r calc_corrplot, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8}
alldata <- alldata %>% mutate(logGDP = log(GDP_USD_2018))

corrplot <- ggpairs(data = alldata, columns = c(5, 15, 8, 10:14),
                    title = "Correlation Matrix, Continuous Variables")
corrplot
```

Strong positive linear relationships are seen between `HDIindex` and `SPI`, `happiness`, and `birth_MF`; between `SPI` and `happiness`, `birth_MF`, and `sixty_MF`; and between `happiness` and `sixty_MF`.  Additionally, strong positive relationships that are possibly nonlinear are seen between `HDI_index` and `sixty_MF`, and between `birth_MF` and `sixty_MF`.

Strong negative relationships are seen between `infantmort` and `birth_MF`, between `HDIindex` and `infantmort`, and between `SPI` and `infantmort`, though the latter two of these may not necessarily be linear.  A strong negative nonlinear relationship is seen between `infantmort` and `sixty_MF`.

Since the goal of this analysis is to compare countries with particular focus on the United States, factor-ordered bivariate plots were generated to explore how the countries compare across the variables of interest, with the United States denoted in red.

First, the top and bottom 20 countries were compared by Social Progress Index:

```{r SPI_country_out, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 6}
SPI_country_point
```

The United States ranks twentieth in social progress.

Next, exploring GDP by country:

```{r GDP_country, echo = FALSE, fig.height = 6, fig.width = 9}
alldata_GDP <- alldata %>%
  filter(!is.na(GDP_USD_2018) == TRUE) %>% 
  arrange(desc(GDP_USD_2018)) %>% 
  select(GDP_USD_2018, country, US, color)

alldata_GDP_top20 <- alldata_GDP %>% head(20)
alldata_GDP_bot20 <- alldata_GDP %>% tail(20)
alldata_GDP_40 <- bind_rows(alldata_GDP_top20, alldata_GDP_bot20)

colors <- alldata_GDP_40$color[order(alldata_GDP_40$GDP_USD_2018)]

GDP_country_point <- ggplot(data = alldata_GDP_40, 
                            aes(x = log(GDP_USD_2018), 
                                y = fct_reorder(country, GDP_USD_2018), 
                                color = US)) +
  geom_point() +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$")) +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlab("log(Gross Domestic Product) ($US 2018)") +
  ylab("Country") +
  ggtitle("Gross Domestic Product by Country, Log Scale, \nTop and Bottom 20 Countries")
GDP_country_point
```

The United States has the world's largest GDP.
 
Next, World Happiness Score:

```{r WHR_country, echo = FALSE, fig.height = 6, fig.width = 6}
alldata_WHR <- alldata %>%
  filter(!is.na(happiness) == TRUE) %>% 
  arrange(desc(happiness)) %>% 
  select(happiness, country, US, color)

alldata_WHR_top20 <- alldata_WHR %>% head(20)
alldata_WHR_bot20 <- alldata_WHR %>% tail(20)
alldata_WHR_40 <- bind_rows(alldata_WHR_top20, alldata_WHR_bot20)

colors <- alldata_WHR_40$color[order(alldata_WHR_40$happiness)]

WHR_country_point <- ggplot(data = alldata_WHR_40, 
                            aes(x = happiness, y = fct_reorder(country, happiness), color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 10) +
  xlab("World Happiness Score") +
  ylab("Country") +
  ggtitle("World Happiness Score by Country, \nTop and Bottom 20 Countries")
WHR_country_point
```

```{r WHR_rank, include = TRUE, echo = TRUE}
which(alldata_WHR$country == "United States")
```

The United States is not among the top 20 countries in terms of happiness; it ranks 21st.

Next, the Human Development Index:

```{r HDI_country, echo = FALSE, fig.height = 6, fig.width = 6}
alldata_HDI <- alldata %>%
  filter(!is.na(HDIindex) == TRUE) %>% 
  arrange(desc(HDIindex)) %>% 
  select(HDIindex, country, US, color)

alldata_HDI_top20 <- alldata_HDI %>% head(20)
alldata_HDI_bot20 <- alldata_HDI %>% tail(20)
alldata_HDI40 <- bind_rows(alldata_HDI_top20, alldata_HDI_bot20)

colors <- alldata_HDI40$color[order(alldata_HDI40$HDIindex)]

HDI_country_point <- ggplot(data = alldata_HDI40, 
                            aes(x = HDIindex, y = fct_reorder(country, HDIindex), color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 1) +
  xlab("HDI") +
  ylab("Country") +
  ggtitle("Human Development Index by Country, \nTop and Bottom 20 Countries")
HDI_country_point
```

The United States ranks twelfth by HDI.

Exploring gender equality:

```{r gender_country, echo = FALSE, fig.height = 6, fig.width = 6}
alldata_gender <- alldata %>%
  filter(!is.na(genderequality_index) == TRUE) %>% 
  arrange(desc(genderequality_index)) %>% 
  select(genderequality_index, country, US, color)

alldata_gender_top20 <- alldata_gender %>% head(20)
alldata_gender_bot20 <- alldata_gender %>% tail(20)
alldata_gender_40 <- bind_rows(alldata_gender_top20, alldata_gender_bot20)

colors <- alldata_gender_40$color[order(alldata_gender_40$genderequality_index)]

gender_country_point <- ggplot(data = alldata_gender_40, 
                            aes(x = genderequality_index, 
                                y = fct_reorder(country, genderequality_index), 
                                color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 1) +
  xlab("Gender Equality Index") +
  ylab("Country") +
  ggtitle("Gender Equality Index by Country, \nTop and Bottom 20 Countries")
gender_country_point
```

```{r gender_rank, include = TRUE, echo = TRUE}
which(alldata_gender$country == "United States")
```

The United States is not among the top 20 countries in terms of gender equality; it ranks 45th.

Examining infant mortality:

```{r infantmort_country, echo = FALSE, fig.height = 6, fig.width = 6}
alldata_infantmort <- alldata %>%
  filter(!is.na(infantmort) == TRUE) %>% 
  arrange(desc(infantmort)) %>% 
  select(infantmort, country, US, color)

alldata_infantmort_top20 <- alldata_infantmort %>% head(20)
alldata_infantmort_bot20 <- alldata_infantmort %>% tail(20)
alldata_infantmort_40 <- bind_rows(alldata_infantmort_top20, alldata_infantmort_bot20)

colors <- alldata_infantmort_40$color[order(alldata_infantmort_40$infantmort)]

infantmort_country_point <- ggplot(data = alldata_infantmort_40, 
                                   aes(x = infantmort, 
                                       y = fct_reorder(country, infantmort), 
                                       color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlab("Infant Mortality Rate") +
  ylab("Country") +
  ggtitle("Infant Mortality Rate, \nTop and Bottom 20 Countries")
infantmort_country_point
```

```{r infantmort_rank, include = TRUE, echo = TRUE}
alldata_infantmort_asc <- alldata_infantmort %>% arrange(infantmort)
which(alldata_infantmort_asc$country == "United States")
```

The United States has the world's 46th lowest infant mortality rate.

Finally, exploring life expectancy:

```{r lifeexp_country, echo = FALSE, fig.height = 6, fig.width = 12}
alldata_lifeexp <- alldata %>%
  filter(!is.na(birth_MF) == TRUE & !is.na(sixty_MF) == TRUE)

alldata_lifeexp_birth <- alldata_lifeexp %>% 
  arrange(desc(birth_MF)) %>% 
  select(birth_MF, country, US, color)

alldata_lifeexp_birth_top20 <- alldata_lifeexp_birth %>% head(20)
alldata_lifeexp_birth_bot20 <- alldata_lifeexp_birth %>% tail(20)
alldata_lifeexp_birth_40 <- bind_rows(alldata_lifeexp_birth_top20, alldata_lifeexp_birth_bot20)

colors <- alldata_lifeexp_birth_40$color[order(alldata_lifeexp_birth_40$birth_MF)]

birth_MF_country_point <- ggplot(data = alldata_lifeexp_birth_40, 
                            aes(x = birth_MF, 
                                y = fct_reorder(country, birth_MF), 
                                color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 100) +
  xlab("Total Life Expectancy (years)") +
  ylab("Country") +
  ggtitle("...At Birth")

alldata_lifeexp_sixty <- alldata_lifeexp %>% 
  arrange(desc(sixty_MF)) %>% 
  select(sixty_MF, country, US, color)

alldata_lifeexp_sixty_top20 <- alldata_lifeexp_sixty %>% head(20)
alldata_lifeexp_sixty_bot20 <- alldata_lifeexp_sixty %>% tail(20)
alldata_lifeexp_sixty_40 <- bind_rows(alldata_lifeexp_sixty_top20, alldata_lifeexp_sixty_bot20)

colors <- alldata_lifeexp_sixty_40$color[order(alldata_lifeexp_sixty_40$sixty_MF)]

sixty_MF_country_point <- ggplot(data = alldata_lifeexp_sixty_40, 
                            aes(x = 60 + sixty_MF, 
                                y = fct_reorder(country, sixty_MF), 
                                color = US)) +
  geom_point() +
  scale_color_manual(values = c("US" = "red", "Non US" = "#65ADC2")) +
  theme(axis.text.y = element_text(color = colors)) +
  guides(color = FALSE) +
  xlim(0, 100) +
  xlab("Total Life Expectancy (years)") +
  ylab("Country") +
  ggtitle("...At Age 60")

grid.arrange(birth_MF_country_point, sixty_MF_country_point, nrow = 1,
             top = textGrob("Total Life Expectancy by Country...",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#545454")))
```

```{r lifeexp_ranks, include = TRUE, echo = TRUE}
which(alldata_lifeexp_birth$country == "United States")
which(alldata_lifeexp_sixty$country == "United States")
```

Once again, the United States is not among the top 20 countries for life expectancy, ranking 34th and 31st respectively for life expectancy at birth and at 60 years of age.


### **Clustering Analysis**

```{r NAomit, include = FALSE, echo = FALSE, eval = TRUE}
clusterdata <- alldata %>%
  na.omit()

# write_csv(clusterdata, paste0("data/clusterdata_", lubridate::today(),".csv"))  # Uncomment to export clustering dataset
```

```{r kmeans_fxn, include = TRUE, echo = TRUE}
kmdf <- function(data, x, y, z){
  kmdata <- data %>% 
    select(x, y, z)
  kmdata <- return(kmdata)
}
```

Clustering happiness versus log GDP:

```{r kmeans_SPI_GDP, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6}
kmdata <- kmdf(clusterdata, "country", "SPI", "logGDP")

set.seed(19811221)
km_SPI_GDP <- kmeans(kmdata[, 2:3], 4)
km_SPI_GDP_cluster <- as.factor(km_SPI_GDP$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_GDP_cluster)

km_SPI_GDP_plot <- ggplot(data = clusterdata1, 
                          aes(x = logGDP, y = SPI, 
                              color = km_SPI_GDP_cluster,
                              size = US,
                              shape = HDI_cat)) +
  geom_point() +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$")) +
  guides(color = FALSE, size = FALSE, shape = guide_legend(reverse = TRUE)) +
  xlab("log(GDP) ($US 2018)") +
  ylab("Social Progress Index") +
  ggtitle("Cluster Analysis, Social Progress Index vs. GDP")
km_SPI_GDP_plot
```

Clustering social progress versus happiness (code for this and subsequent clustering not shown for brevity):

```{r kmeans_SPI_happiness, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6}
kmdata <- kmdf(clusterdata, "country", "SPI", "happiness")

set.seed(19811221)
km_SPI_happiness <- kmeans(kmdata[, 2:3], 4)
km_SPI_happiness_cluster <- as.factor(km_SPI_happiness$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_happiness_cluster)

km_SPI_happiness_plot <- ggplot(data = clusterdata1, 
                                aes(x = happiness, y = SPI, 
                                    color = km_SPI_happiness_cluster,
                                    size = US,
                                    shape = HDI_cat)) +
  geom_point() +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  guides(color = FALSE, size = FALSE, shape = guide_legend(reverse = TRUE)) +
  xlab("World Happiness Score") +
  ylab("Social Progress Index") +
  ggtitle("Cluster Analysis, Social Progress Index vs. World Happiness Score")
km_SPI_happiness_plot
```

Clustering social progress versus gender equality:

```{r kmeans_SPI_gender, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6}
kmdata <- kmdf(clusterdata, "country", "SPI", "genderequality_index")

set.seed(19811221)
km_SPI_gender <- kmeans(kmdata[, 2:3], 4)
km_SPI_gender_cluster <- as.factor(km_SPI_gender$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_gender_cluster)

km_SPI_gender_plot <- ggplot(data = clusterdata1, 
                                aes(x = genderequality_index, y = SPI, 
                                    color = km_SPI_gender_cluster,
                                    size = US,
                                    shape = HDI_cat)) +
  geom_point() +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  guides(color = FALSE, size = FALSE, shape = guide_legend(reverse = TRUE)) +
  xlab("Gender Equality Index") +
  ylab("Social Progress Index") +
  ggtitle("Cluster Analysis, Social Progress Index vs. Gender Equality Index")
km_SPI_gender_plot
```

Clustering social progress versus infant mortality:

```{r kmeans_SPI_infantmort, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6}
kmdata <- kmdf(clusterdata, "country", "SPI", "infantmort")

set.seed(19811221)
km_SPI_infantmort <- kmeans(kmdata[, 2:3], 4)
km_SPI_infantmort_cluster <- as.factor(km_SPI_infantmort$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_infantmort_cluster)

km_SPI_infantmort_plot <- ggplot(data = clusterdata1, 
                                aes(x = infantmort, y = SPI, 
                                    color = km_SPI_infantmort_cluster,
                                    size = US,
                                    shape = HDI_cat)) +
  geom_point() +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  guides(color = FALSE, size = FALSE, shape = guide_legend(reverse = TRUE)) +
  xlab("Infant Mortality Rate") +
  ylab("Social Progress Index") +
  ggtitle("Cluster Analysis, Social Progress Index vs. Infant Mortality Rate")
km_SPI_infantmort_plot
```

Clustering social progress versus life expectancy:

```{r kmeans_SPI_lifeexp, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6}
kmdata <- kmdf(clusterdata, "country", "SPI", "birth_MF")

set.seed(19811221)
km_SPI_birth_MF <- kmeans(kmdata[, 2:3], 4)
km_SPI_birth_MF_cluster <- as.factor(km_SPI_birth_MF$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_birth_MF_cluster)

km_SPI_birth_MF_plot <- ggplot(data = clusterdata1, 
                               aes(x = birth_MF, y = SPI, 
                                   color = km_SPI_birth_MF_cluster,
                                   size = US,
                                   shape = HDI_cat)) +
  geom_point() +
  xlim(50, 90) +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  guides(color = FALSE, size = FALSE, shape = FALSE) +
  xlab("Total Life Expectancy at Birth (years)") +
  ylab("Social Progress Index") +
  ggtitle("...at Birth")

kmdata <- kmdf(clusterdata, "country", "SPI", "sixty_MF")

set.seed(19811221)
km_SPI_sixty_MF <- kmeans(kmdata[, 2:3], 4)
km_SPI_sixty_MF_cluster <- as.factor(km_SPI_sixty_MF$cluster)

clusterdata1 <- cbind(clusterdata, km_SPI_sixty_MF_cluster)

km_SPI_sixty_MF_plot <- ggplot(data = clusterdata1, 
                                aes(x = 60 + sixty_MF, y = SPI, 
                                    color = km_SPI_sixty_MF_cluster,
                                    size = US,
                                    shape = HDI_cat)) +
  geom_point() +
  xlim(50, 90) +
  ylim(0, 100) +
  scale_shape_manual(values = c(18, 17, 15, 16)) +
  guides(color = FALSE, size = FALSE, shape = guide_legend(reverse = TRUE, nrow = 1)) +
  theme(legend.position = "bottom") +
  xlab("Total Life Expectancy at Age Sixty (years)") +
  ylab("Social Progress Index") +
  ggtitle("...at Age 60")

grid.arrange(km_SPI_birth_MF_plot, km_SPI_sixty_MF_plot, nrow = 2,
             top = textGrob("Cluster Analysis, Social Progress Index vs. Total Life Expectancy...",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#545454")))
```



## **Discussion**



### *Limitations*



## **Conclusion**



## **References**